---
title: "PCOS project"
author: "Praveen Sinha"
date: "2026-04-12"
output: html_document
---

### load and prepare data
```{r}
df <- read.csv("../../data/PCOS_data.csv")
```

### Clean column names
```{r}
colnames(df) <- make.names(colnames(df))
```

### Create target
```{r}
df$PCOS <- df$PCOS..Y.N.
```

### Drop unwanted columns
```{r}
df <- df[, !(names(df) %in% c(
  "Sl..No", 
  "Patient.File.No.", 
  "PCOS..Y.N.", 
  "X"
))]
```

### Convert all columns to numeric 
```{r}
df[] <- lapply(df, function(col) {
  if (is.factor(col) || is.character(col)) {
    as.numeric(as.factor(col))
  } else {
    col
  }
})
```

### Remove rows with NA
```{r}
df <- na.omit(df)
```
### Split features and target
```{r}
X <- as.matrix(df[, !(names(df) %in% c("PCOS"))])
y <- df$PCOS
```

### Add intercept
```{r}
X <- cbind(1, X)
```

```{r}
str(X)
```

### Scale features
```{r}
X[, -1] <- scale(X[, -1])

## defining the metropolis-hastings and helper functions
sigmoid <- function(z) {
  1 / (1 + exp(-z))
}

log_likelihood <- function(w, X, y) {
  p <- sigmoid(X %*% w)
  sum(y * log(p + 1e-9) + (1 - y) * log(1 - p + 1e-9))
}

log_prior <- function(w) {
  sum(dnorm(w, mean = 0, sd = 1, log = TRUE))
}

log_posterior <- function(w, X, y) {
  log_likelihood(w, X, y) + log_prior(w)
}

metropolis_hastings <- function(X, y, n_iter = 5000, step_size = 0.1) {
  
  n_features <- ncol(X)
  w_current <- rep(0, n_features)
  
  samples <- matrix(0, nrow = n_iter, ncol = n_features)
  
  accept_count <- 0
  
  for (i in 1:n_iter) {
    
    w_proposal <- w_current + rnorm(n_features, 0, step_size)
    
    log_post_current <- log_posterior(w_current, X, y)
    log_post_proposal <- log_posterior(w_proposal, X, y)
    
    log_alpha <- log_post_proposal - log_post_current
    
    if (log(runif(1)) < log_alpha) {
      w_current <- w_proposal
      accept_count <- accept_count + 1
    }
    
    samples[i, ] <- w_current
  }
  
  cat("Acceptance Rate:", accept_count / n_iter, "\n")
  
  return(samples)
}
```

### Sample
```{r}
samples <- metropolis_hastings(X, y, n_iter = 5000, step_size = 0.05)


burn_in <- 1000
posterior <- samples[(burn_in+1):nrow(samples), ]

# smaller step size
samples2 <- metropolis_hastings(X, y, n_iter = 5000, step_size = 0.03)

# larger step size
samples3 <- metropolis_hastings(X, y, n_iter = 5000, step_size = 0.07)
```

We evaluated multiple step sizes to assess the sampler performance. As we expected, smaller step sizes resulted in higher acceptance rates (≈0.59), indicating overly conservative proposals, while larger step sizes reduced acceptance (≈0.25), reflecting more aggressive exploration. We selected a step size of 0.05, yielding an acceptance rate of approximately 0.41, which provided a strong balance between exploration and stability. The consistency of posterior estimates across step sizes further supports the robustness of the model.

### stable results
```{r}
colMeans(samples[(1000+1):5000, ])
colMeans(samples2[(1000+1):5000, ])
colMeans(samples3[(1000+1):5000, ])
```
### Mean estimates
```{r}
coef_mean <- colMeans(posterior)

# 95% credible intervals
lower <- apply(posterior, 2, quantile, 0.025)
upper <- apply(posterior, 2, quantile, 0.975)

results <- data.frame(
  Feature = colnames(X),
  Mean = coef_mean,
  Lower = lower,
  Upper = upper
)

# Fix intercept name
results$Feature[1] <- "Intercept"

head(results)

important <- results[
  (results$Lower > 0) | (results$Upper < 0),
]

important

library(ggplot2)

ggplot(important[-1, ], aes(x = Mean, y = reorder(Feature, Mean))) +
  geom_point() +
  geom_errorbar(
    aes(xmin = Lower, xmax = Upper),
    orientation = "y",
    width = 0.2
  ) +
  labs(
    title = "Significant Features (95% Credible Intervals)",
    x = "Effect on PCOS",
    y = "Features"
  ) +
  theme_minimal()
```

### posterior distribution plots
```{r} 
hist(posterior[, 10], main="Posterior for Cycle.R.I.", col="lightblue")
```

### Trace plot (for convergence)
```{r} 
plot(samples[,10], type="l", main="Trace Plot")
```

### Coefficients of features
```{r} 

for (i in 1:nrow(results)) {
  cat(
    results$Feature[i], ":\n",
    "  Mean =", round(results$Mean[i], 3), "\n",
    "  95% CI = [",
    round(results$Lower[i], 3), ",",
    round(results$Upper[i], 3), "]\n\n"
  )
}
```

The strongest predictor in our model was follicle count, particularly in the right ovary, which is directly aligned with the Rotterdam criteria for PCOS diagnosis.

### all results
```{r}
write.csv(results, "bayesian_manual_coefficients.csv", row.names = FALSE)
```

### important
```{r}
write.csv(important, "bayesian_important_features.csv", row.names = FALSE)
write.csv(results, "~/Desktop/DS 4420/bayesian_manual_coefficients.csv", row.names = FALSE)
```

### train/test split
```{r}
set.seed(42)

idx <- sample(1:nrow(X), 0.8 * nrow(X))

X_train <- X[idx, ]
y_train <- y[idx]

X_test <- X[-idx, ]
y_test <- y[-idx]
```

### Posterior averaging
```{r}
samples <- metropolis_hastings(X_train, y_train, n_iter = 5000, step_size = 0.05)

burn_in <- 1000
posterior <- samples[(burn_in+1):nrow(samples), ]
```


```{r}
predict_proba <- function(X_new, posterior_samples) {
  probs <- apply(posterior_samples, 1, function(w) {
    1 / (1 + exp(-X_new %*% w))
  })
  
  rowMeans(probs)
}
```

### train accuracy
```{r}
train_probs <- predict_proba(X_train, posterior)
train_preds <- ifelse(train_probs > 0.5, 1, 0)

train_accuracy <- mean(train_preds == y_train)

train_accuracy
```

### test accuracy
```{r}
test_probs <- predict_proba(X_test, posterior)
test_preds <- ifelse(test_probs > 0.5, 1, 0)

test_accuracy <- mean(test_preds == y_test)

test_accuracy
```

### confusion matrix
```{r}
table(Predicted = test_preds, Actual = y_test)
```

### Precision + Recall
```{r}
precision <- sum(test_preds == 1 & y_test == 1) / sum(test_preds == 1)
recall <- sum(test_preds == 1 & y_test == 1) / sum(y_test == 1)

precision
recall
```